---
title: "Practical Machine Learning  Project 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<h1>Practical Machine Learning - Course Project</h1>
<h3> Silvester Makgae
September 25, 2018</h3>
<h2>Practical Machine Learning Project : Prediction Assignment Writeup</h2>
<b>I. Overview</b>
<p>This document is the final report of the Peer Assessment project from Coursera's course
Practical Machine Learning, as part of the Specialization in Data Science. It was built up in
RStudio, using its knitr functions, meant to be published in html format.</p>
<p>This analysis meant to be the basis for the course quiz and a prediction assignment writeup.
The main goal of the project is to predict the manner in which 6 participants performed some
exercise as described below. This is the "classe" variable in the training set. The machine
learning algorithm described here is applied to the 20 test cases available in the test data and
the predictions are submitted in appropriate format to the Course Project Prediction Quiz for
automated grading.</p>
<b>II. Background</b>
<p>Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a
large amount of data about personal activity relatively inexpensively. These type of devices are
part of the quantified self movement - a group of enthusiasts who take measurements about
themselves regularly to improve their health, to find patterns in their behavior, or because they
are tech geeks. One thing that people regularly do is quantify how much of a particular activity
they do, but they rarely quantify how well they do it. In this project, your goal will be to use data
from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked
to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available
from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight
Lifting Exercise Dataset).
Read more: http://groupware.les.inf.puc-rio.br/har#ixzz3xsbS5bVX</p>
<b>III. Data Loading and Exploratory Analysis</b>
<b>a) Dataset Overview</b>
<p>The training data for this project are available here:<br>
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv<br>
The test data are available here:<br>
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv</p>
<p>The data for this project come from http://groupware.les.inf.puc-rio.br/har. Full source:
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. "Qualitative Activity Recognition of
Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with
SIGCHI (Augmented Human '13)". Stuttgart, Germany: ACM SIGCHI, 2013.<br>
My special thanks to the above mentioned authors for being so generous in allowing their data
to be used for this kind of assignment.<br>
A short description of the datasets content from the authors' website:<br>
"Six young health participants were asked to perform one set of 10 repetitions of the Unilateral
Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A),
throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering
the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).
Class A corresponds to the specified execution of the exercise, while the other 4 classes
correspond to common mistakes. Participants were supervised by an experienced weight lifter
to make sure the execution complied to the manner they were supposed to simulate. The
exercises were performed by six male participants aged between 20-28 years, with little weight
lifting experience. We made sure that all participants could easily simulate the mistakes in a
safe and controlled manner by using a relatively light dumbbell (1.25kg)."</p>
<b>b) Environment Preparation</b>
<p>We first upload the R libraries that are necessary for the complete analysis.</p>
```{r}
rm(list=ls()) # free up memory for the download of the data sets
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(12345)
```
<p>
<b>c) Data Loading and Cleaning</b>
The next step is loading the dataset from the URL provided above. The training dataset is then
partinioned in 2 to create a Training set (70% of the data) for the modeling process and a Test
set (with the remaining 30%) for the validations. The testing dataset is not changed and will only
be used for the quiz results generation.</p>
```{r}
# set the URL for the download
UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```

```{r}
# download the datasets
training <- read.csv(url(UrlTrain))
testing <- read.csv(url(UrlTest))
```

```{r}
# create a partition with the training dataset
inTrain <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet <- training[-inTrain, ]
dim(TrainSet)
dim(TestSet)
```

<p>Both created datasets have 160 variables. Those variables have plenty of NA, that can be
removed with the cleaning procedures below. The Near Zero variance (NZV) variables are also
removed and the ID variables as well.</p>

```{r}
# remove variables with Nearly Zero Variance
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet <- TestSet[, -NZV]
dim(TrainSet)
```
```{r}
dim(TestSet)
```

```{r}
# remove variables that are mostly NA
AllNA <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet <- TestSet[, AllNA==FALSE]
dim(TrainSet)
```

```{r}
dim(TestSet)
```

```{r}
# remove identification only variables (columns 1 to 5)
TrainSet <- TrainSet[, -(1:5)]
TestSet <- TestSet[, -(1:5)]
dim(TrainSet)
```

```{r}
dim(TestSet)
```
<p>
With the cleaning process above, the number of variables for the analysis has been reduced to
54 only.<br>
<b>d) Correlation Analysis</b><br>
A correlation among variables is analysed before proceeding to the modeling procedures.</p>
```{r}
corMatrix <- cor(TrainSet[, -54])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower",
tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

<p>The highly correlated variables are shown in dark colors in the graph above. To make an evem
more compact analysis, a PCA (Principal Components Analysis) could be performed as
pre-processing step to the datasets. Nevertheless, as the correlations are quite few, this step
will not be applied for this assignment.<br>
<b>IV. Prediction Model Building</b><br>
Three methods will be applied to model the regressions (in the Train dataset) and the best one
(with higher accuracy when applied to the Test dataset) will be used for the quiz predictions.
The methods are: Random Forests, Decision Tree and Generalized Boosted Model, as
described below.<br>
A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the
models.<br>
<b>a) Method: Random Forest</b><br>
```{r}
# model fit
set.seed(12345)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf",
trControl=controlRF)
modFitRandForest$finalModel
# prediction on Test dataset
predictRandForest <- predict(modFitRandForest, newdata=TestSet)
confMatRandForest <- confusionMatrix(predictRandForest, TestSet$classe)
confMatRandForest
# plot matrix results
plot(confMatRandForest$table, col = confMatRandForest$byClass,
main = paste("Random Forest - Accuracy =",
round(confMatRandForest$overall['Accuracy'], 4)))
```


<b>b) Method: Decision Trees</b><br>
```{r}
# model fit
set.seed(12345)
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modFitDecTree)
```

```{r}
# prediction on Test dataset
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatDecTree
```

```{r}
# plot matrix results
plot(confMatDecTree$table, col = confMatDecTree$byClass,
main = paste("Decision Tree - Accuracy =",
round(confMatDecTree$overall['Accuracy'], 4)))
```

<b>c) Method: Generalized Boosted Model</b><br>
```{r}
# model fit
set.seed(12345)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM <- train(classe ~ ., data=TrainSet, method = "gbm",
trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
```
```{r}
## A gradient boosted model with multinomial loss function.
## 150 iterations were performed.
## There were 53 predictors of which 41 had non-zero influence.
# prediction on Test dataset
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, TestSet$classe)
confMatGBM
```
```{r}
# plot matrix results
plot(confMatGBM$table, col = confMatGBM$byClass,
main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))
```


<p><b>V. Applying the Selected Model to the Test Data<b><p>
The accuracy of the 3 regression modeling methods above are:<br>
Random Forest : 0.9963<br>
Decision Tree : 0.7368<br>
GBM : 0.9839<br>
In that case, the Random Forest model will be applied to predict the 20 quiz results (testing
dataset) as shown below.<br></p>
```{r}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```

